Memes: an R interface to the MEME Suite

### 2020-09-22 manuscript.Rmd
Draft complete.

---
title: '{memes}: an R interface to the MEME Suite'
author:
- ORCID: null
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: 
 - word_document:
   fig.caption: yes
---

```{r setup, echo = FALSE}
# Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
suppressPackageStartupMessages(library(memes))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(ggplot2))
drake::r_make()
```

@@

The MEME Suite family of tools provides a
comprehensive set of utilities to interrogate motif content across a broad range
of biological contexts. However, several factors limit the full potential of
these tools for use in data analysis. 
  1) must do significant preprocessing to get data in correct input format
    - written to analyze 1 group of sequences, not amenable to multigroup designs
    - no common patterns for generating inputs from ChIP-seq whose data are primarily range based
  2) Easily parseable output data often doesn't contain all information
    - ex tomtom alternative hits in .tsv output only present in IUPAC format, not matrix
      - matrix is present in XML file, but no tools exist to parse this
  3) limited data visualization capabilities
  4) difficult to integrate data with additional sources

Here we present `memes`, a motif analysis package which provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor
data structures provide a full-featured representaiton of MEME Suite output
data, providing users quick access to all relevant data structures with simple
syntax.

@@

The core functionality of `memes` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `run` function
(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
accept sequence and motif inputs as needed and any valid commandline flags as
additional function parameters.

**Describe each tool briefly?**
 - This could also be relegated to a table

**table of comparison to commandline software?**
  - input type:
  - output type:

@@

 - **NOTE: Biopython has a MEME/MAST parser but it works on tsv https://biopython.org/docs/1.74/api/Bio.motifs.meme.html**
 - **Actually, they merged XML import in March: https://github.com/biopython/biopython/pull/2354**
 - **Need to revise above statement, but note that most tools don't have support**

@@

# RETURN
Motif analysis is important for assays like ChIP-seq which store data as genomic
coordinates, not sequence. This is in conflict with MEME Suite which is designed
to study sequences. Therefore, users must write custom code to extract DNA
sequence of their ranges of interest. Although tools like `bedtools getfasta`
can simplify the process of extracting DNA sequence from genomic coordinates,
some MEME tools require fasta headers to be specially formatted to extract full
functionality, requiring users to write additional custom code to properly
format input data.

@@

 - `view_motifs()` plots motif as stack (if I decide to add this feature... need to think about this)
 - `ame_plot_heatmap` plot clustered heatmap of AME results 
   - `ame_compare_heatmap_methods` compare nonlinearity of data visualization in heatmap form
 - `view_tomtom_hits` visualize and compare TomTom matches to assess match quality
   - `force_best_match` can be used to update "best" match demonstrating how data visualization can be useful for data curation

The AME tool searches for enrichment of motifs within groups of sequences
relative to a control set. The meaningful result from this tool is the
statistical parameter (for example, a p-value) associated with the significance
of motif enrichment. However, AME does not provide a mechanism for visualizing
these results. Additionally, AME is often used 
## RETURN

@@

```{r}
dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
```
Using this approach, the results can be visualized using 
the `view_tomtom_hits()` function to visually inspect the matches
assigned by TomTom, providing a simple way to assess the quality of the match. 
A representative plot of these data is shown in \ref{fig:tomtom_example}, while
the full results can be viewed in \ref{tab:tomtom_table}. We
uncover that the motifs associated with decreasing sites are highly similar to
the E93 motif. Conversely, motifs found in increasing sites match the
transcription factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These
data support the hypothesis that E93 binding to its motif may play a larger role
in chromatin closing activity, while binding of the transcription factor br is
associated with chromatin opening.
```{r tomtom_example, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

@@

dreme_vs_static_tomtom %>% 
  purrr::imap(~{
    .x %>% 
      dplyr::mutate(name = paste0(.y, " ", name)) %>% 
      update_motifs() %>% 
      plot_top_hit() %>% 
      .[[2]]
    }) %>% 
  patchwork::wrap_plots() +
    patchwork::plot_annotation(tag_levels = "A")
```
```{r}
library(gt)

dreme_vs_static_tomtom %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste(category, "Motifs")) %>% 
  dplyr::arrange(category, rank) %>% 
  dplyr::select(category, name, best_match_altname) %>% 
  dplyr::rename("Best Match TF" = best_match_altname) %>% 
  gt(rowname_col = "name",
     groupname_col = "category",
     lab = "tab:tomtom_table")

>>>>

---
title: 'Memes: an R interface to the MEME Suite'
author:
- ORCID: 0000-0003-1000-1579
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: 
 - word_document
---

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
# Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
suppressPackageStartupMessages(library(memes))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ComplexHeatmap))
drake::r_make()
```

@@

The MEME Suite family of tools provides a comprehensive set of utilities to
interrogate motif content across a broad range of biological contexts. However,
several factors limit the full potential of these tools for use in data
analysis. MEME Suite tools require carefully formatted inputs to achieve full
functionality, yet few tools exist to simplify the process of data formatting,
requiring instead that users write custom code to prepare their data.
Furthermore, the output data from each MEME Suite tool often have complex
structures which must be parsed to extract the full suite of information, again
requiring users to write custom code for this task. Finally, although the
data-generation capabilities of the MEME Suite are excellent, the tools lack
powerful ways to visualize the results. These factors act as barriers to
adoption and preclude deep analysis of the data.

Here we present `memes`, a motif analysis package which provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor
data structures provide a full-featured representation of MEME Suite output
data, providing users quick access to all relevant data structures with simple
syntax.

@@

The core functionality of `memes` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `run` function
(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
accept sequence and motif inputs as needed and any valid commandline flags as
additional function parameters. A full list of currently supported MEME Suite
tools and their function is found in \ref{tab:tool_function}.

@@

 - **NOTE: Biopython has a MEME/MAST parser but it works on tsv https://biopython.org/docs/1.74/api/Bio.motifs.meme.html**
 - **Actually, they merged XML import in March: https://github.com/biopython/biopython/pull/2354**
 - **Need to revise above statement, but note that most tools don't have support**
 - ? To our knowledge, no publicly available tools exist to parse XML formatted data from all MEME Suite tools ...

@@

Motif analysis is important for assays like ChIP-seq which store data as genomic
coordinates, not sequence. This is in conflict with MEME Suite which is designed
to study sequences. Therefore, users must write custom code to extract DNA
sequence of their ranges of interest. Although tools like `bedtools getfasta`
can simplify the process of extracting DNA sequence from genomic coordinates,
some MEME tools require fasta headers to be specially formatted to extract full
functionality, requiring users to write additional custom code to properly
format input data.

@@

The TomTom tool allows users to compare unknown motifs to a set of known motifs
to identify a best match. Visual inspection of motif comparison data is a key
step in assessing the quality and correctness of a match. The `view_tomtom_hits`
function allows users to compare query motifs with the list of potential matches
as assigned by TomTom. Often, matching tools are useful for reducing the
search-space for potential candidates, but users are free to leverage their
understanding of the biology to override these matches. The `force_best_match`
function allows users to quickly reassign the TomTom best match to a
lower-ranked, but still statistically significant match to ease downstream data
processing.

The AME tool searches for enrichment of motifs within groups of sequences
relative to a control set. The meaningful result from this tool is the
statistical parameter (for example, a p-value) associated with the significance
of motif enrichment. However, AME does not provide a mechanism for visualizing
these results. Additionally, AME is often used to examine motif content of
multiple groups of sequences, therefore a plot which can represent statistical
significance and group membership of a motif match could be a useful way of
visualizing the data. `memes` visualized AME results with the `ame_plot_heatmap`
function, which returns a `ggplot2` formatted heatmap comparing motif
significance within one or multiple groups. Several options exist to customize
the heatmap values to help capture different aspects of the output data. The
`ame_compare_heatmap_methods()` function enables users to compare the
distribution of values between samples in order to select a threshold which
accurately captures the dynamic range of their results.

@@

```{r}
dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
```
Using this approach, the results can be visualized using 
the `view_tomtom_hits()` function to visually inspect the matches
assigned by TomTom, providing a simple way to assess the quality of the match. 
A representative plot of these data is shown in \ref{fig:tomtom_example}, while
the full results can be viewed in \ref{tab:tomtom_table}. We
uncover that the motifs associated with decreasing sites are highly similar to
the E93 motif. Conversely, motifs found in increasing sites match the
transcription factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These
data support the hypothesis that E93 binding to its motif may play a larger role
in chromatin closing activity, while binding of the transcription factor br is
associated with chromatin opening.
```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

@@

dreme_vs_static_tomtom %>% 
  purrr::imap(~{
    .x %>% 
      dplyr::mutate(name = paste0(.y, " ", name)) %>% 
      update_motifs() %>% 
      plot_top_hit() %>% 
      .[[2]]
    }) %>% 
  patchwork::wrap_plots() +
    patchwork::plot_annotation(tag_levels = "A")
```
```{r, echo=F}
library(gt)

dreme_vs_static_tomtom %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste(category, "Motifs")) %>% 
  dplyr::arrange(category, rank) %>% 
  dplyr::select(category, name, best_match_altname) %>% 
  dplyr::rename("Best Match TF" = best_match_altname) %>% 
  gt(rowname_col = "name",
     groupname_col = "category")
     # this feature is in development branch??
     # search SO: gt table reference
     #lab = "tab:tomtom_table")
